Optimistic Atomic Multicast in One Communication Delay

### 2011-07-27 textos/optpaxos/draft_dropbox.tex
Optimized atomic multicast

Hereafter, we assume that in addition to a set of destination groups, $m.dst$, each message $m$ has a source group, $m.src$, and a timestamp, $m.ts$. To \amcast{} $m$, process $p$ in group $g$ sets $m.ts$ to a unique timestamp based on its real-time local clock and then \rmcast{}s $m$ to all processes in $g$.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We use a $getTime()$ primitive, which returns the current value of the local wallclock, which is monotonically increasing. We also use the concept of \textit{barrier}. A barrier is the certainty that a given group will not send any new messages with a timestamp lower than a given value to some other group. As all groups receive messages from each other in the order of their timestamps, which is possible by using \rmcast{}, every message is seen as a barrier.

@@

Even though each message $m$ might not have its source group as a destination, $m$ is still agreed upon in its group of origin, so its order among other messages from $m.src$ is decided and it %$m.gs$ can be decided, and also for 
may be retrievable even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

Whenever $p$ has some message $m$ to \amcast, it assigns to $m$ an initial timestamp value, based on the local wallclock of $p$, after which $m$ is \rmcast{} to all the other processes of $g$ (l. \ref{algline:gettime} and l. \ref{algline:rmlocal}). After \rmdel{}ing $m$, each process puts it in $\localmsgs$ to be proposed in $g$ until it is decided (l. \ref{algline:undecided} and l. \ref{algline:propose}), along with other undecided messages. As all correct processes of $g$ do so, some accepted proposal will end up containing $m$.

@@

Let $h$ be any possible destination group of $m$ other than $g$. When some process $q$ of $h$ \rmdel{}s $m$ for the first time, $q$ sets $barrier(g)$ to $m.ts$ (l. \ref{algline:incbar}), since, from the \rmcast{} primitive, any previous message from $g$ has already been received. Besides, $q$ knows that the processes of $g$ have already decided the final timestamp of $m$, so $q$ inserts it into $\stamped$ (l. \ref{algline:insbp1}).

For any process $r$, of any group $i$ in $m.dst$, when $m$ belongs to $\stamped$, having the lowest timestamp among all undelivered messages of such set (l. \ref{algline:condel1}), and $r$ has already received a barrier from every group in $sendersTo(i)$, where the value of such barrier is greater than $m.ts$ (l. \ref{algline:condel2}), $m$ is \amdel{}ed by $r$ (l. \ref{algline:consdeliver}). Doing so, no message $m'$ with a timestamp greater than $m.ts$ will be \amdel{}ed before $m$ by any process in $m.dst$.

The consensus instances within $g$ are decided in the same order for every correct process of $g$. Such order is used to define the delivery order of $g$'s messages as well. We want this order to be consistent with the timestamps of the messages, so such timestamps can be changed if they are not in accordance with the delivery order.

@@

To allow for the timestamp of messages to be increased and still have these messages delivered as soon as possible, their timestamps are increased by an infinitesimal value. For that reason, each timestamp value will consist of a real-time clock value, and a sequence value, which is used only when messages need to have their timestamps changed. Therefore, we have that $m.ts = (rtc, seq)$, where $rtc$ is some value related to real-time clock of a process and $seq$ is a sequence number to define an order between messages with the same $rtc$. Then we have:

\begin{align*}
m.ts < m'.ts \Longleftrightarrow m.ts.rtc < m'.ts.rtc \vee (m.ts.rtc = m'.ts.rtc \wedge m.ts.seq < m'.ts.seq)\\
%&\vee (m.ts.rtc = m'.ts.rtc \wedge m.ts.seq = m'.ts.seq \wedge m.ts.p_{id} < m'.ts.p_{id})
\end{align*}

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $\localmsgs \leftarrow \emptyset$, $\decided \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{To} \amcast{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$ \label{algline:gettime}
  \STATE \rmcastarg{$\{g\}$}{$m$} \label{algline:rmlocal}
\ENDINDENT

\blankline
\WHEN{$\localmsgs \setminus \decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
	\STATE $undecided \leftarrow \localmsgs \setminus \decided$ \label{algline:undecided}
	\STATE propose$_g$($k, undecided$) \label{algline:propose}
	\STATE \textbf{wait until} decide$_g$($k, msgSet$) \label{algline:decide}
	\WHILE{$msgSet \setminus \decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus \decided$ with smallest timestamp \label{algline:decidedloopbegin}
		\STATE let $m'$ be the message in $\decided$ with greatest timestamp
		\IF{$m'$ exists and $m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:changetsafterdecision}
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $\stamped \leftarrow \stamped \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $\decided \leftarrow \decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE $barrier(g) \leftarrow m.ts$ \label{algline:incbarlocal}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE

@@

\blankline
\INDENT{\textbf{when} deciding a message $m$}
  \INDENT{\textbf{for all} $h \in m.dst \cup \{g\}$ \textbf{do}}
    \STATE $lastBarrierCreated(h) \leftarrow m.ts.rtc$ 
  \ENDINDENT  
\ENDINDENT

\blankline
\INDENT{\textbf{when} $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $\localmsgs \leftarrow \localmsgs \cup \{null\}$ \label{algline:enqnull}
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

The problem with addressing liveness this way is that, in the worst case, $G'$ has decided a message $m$ and has just received the last barrier $b$ from $G$, such that $b < m.ts$. This would mean that, if $G$ has no messages to send to $G'$, $G'$ will have to wait, at least, for $barrierThreshold$ -- maybe just to receive some $b' : b' < m.ts$, having to wait again and so on. How long exactly it will take to \cons{} $m$ depends on many variables, such as how far in the past $m$ was created and how long it takes for some barrier $b: b > m.ts$ to arrive.% Because of that, the time between \amcast{}ing and \cons{}ing $m$ be, in the worst case, $2w(p) + 4\tconsm + barrierThreshold$, where $w(p)$ is the wait window of the process $p \in G$, and \tcons\ is counted twice because there are two consensus proposals -- first in $G'$ regarding $m$, and then in $G$ regarding the empty message.

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $\localmsgs$ set only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such set and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

\subsubsection{Requesting empty messages}
\label{sec:nullondemand}

@@

 \subsection{Performance analysis}

@@

\section{Optimistic atomic multicast}
\label{sec:optdel}

>>>>

Hereafter, we assume that in addition to application data, each message $m$ has four fields: its set of destination groups $m.dst$, its source group $m.src$, its timestamp $m.ts$ and an unique message identifier $m.id$. To \amcast{} $m$, process $p$ in group $g$ sets $m.ts$ to a unique timestamp based on its real-time local clock and then \rmcast{}s $m$ to all processes in $g$.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We use a $getTime()$ primitive, which returns the current value of the local wallclock, which is monotonically increasing. We also use the concept of \textit{barrier}. A barrier is the certainty that a given group will not send any new messages with a timestamp lower than a given value to some other group. As all groups receive messages from each other in the order of their timestamps, which is done by using \rmcast{}, every message is seen as a barrier.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $\localmsgs \leftarrow \emptyset$, $\decided \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow 0$ 
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{To} \amcast{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow getTime()$ \label{algline:gettime}
  \STATE \rmcastarg{$\{g\}$}{$m$} \label{algline:rmlocal}
\ENDINDENT

\blankline
\WHEN{$\localmsgs \setminus \decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
  \STATE $undecided \leftarrow \localmsgs \setminus \decided$ \label{algline:undecided}
  \STATE propose$_g$($k, undecided$) \label{algline:propose}
  \STATE \textbf{wait until} decide$_g$($k, msgSet$) \label{algline:decide}
  \WHILE{$msgSet \setminus \decided \neq \emptyset$}
    \STATE let $m$ be the message in $msgSet \setminus \decided$ with smallest timestamp \label{algline:decidedloopbegin}
    \STATE let $m'$ be the message in $\decided$ with greatest timestamp
    \IF{$m'$ exists and $m'.ts > m.ts$}
      \STATE $m.ts \leftarrow m'.ts + 1$ \label{algline:changetsafterdecision}
    \ENDIF     
    \IF{$g \in m.dst$} \label{algline:checkcons}
      \STATE $\stamped \leftarrow \stamped \cup \{m\}$  \label{algline:insbp2}
    \ENDIF
    \STATE $\decided \leftarrow \decided \cup \{m\}$  \label{algline:addtodecided}
    \STATE $barrier(g) \leftarrow m.ts$ \label{algline:incbarlocal}
    \STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
  \ENDWHILE

% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

Whenever $p$ has some message $m$ to \amcast, it assigns to $m$ an initial timestamp value, based on the local wallclock of $p$, after which $m$ is \rmcast{} to all the other processes of $g$ (l. \ref{algline:gettime} and l. \ref{algline:rmlocal}). After \rmdel{}ing $m$, each process puts it in $\localmsgs$ to be proposed in $g$ until it is decided (l. \ref{algline:undecided} and l. \ref{algline:propose}), along with other undecided messages. As all correct processes of $g$ do so, some accepted proposal will end up containing $m$. Even though $m$ might not have its source group as a destination, $m$ is still agreed upon in its group of origin, so its order among other messages from $g$ is decided and it may be retrievable even in the presence of failures.

@@

Let $h$ be any possible destination group of $m$ other than $g$. When some process $q$ of $h$ \rmdel{}s $m$ for the first time -- $m$ may be \rmdel{}ed multiple times, since all correct processes of $g$ \rmcast{} $m$ to $h$ --, $q$ sets $barrier(g)$ to $m.ts$ (l. \ref{algline:incbar}), since, from the \rmcast{} primitive, any previous message from $g$ has already been received. Besides, $q$ knows that the processes of $g$ have already decided the final timestamp of $m$, so $q$ inserts it into $\stamped$ (l. \ref{algline:insbp1}).

For any process $r$, of any group $i$ in $m.dst$, when $m$ belongs to $\stamped$, having the lowest timestamp among all undelivered messages of such set (l. \ref{algline:condel1}), and $r$ has already received a barrier from every group in $sendersTo(i)$, where the value of such barrier is greater than $m.ts$ (l. \ref{algline:condel2}), $m$ is \amdel{}ed by $r$ (l. \ref{algline:consdeliver}). Doing so, no message $m'$ with a timestamp greater than $m.ts$ will be \amdel{}ed before $m$ by process $r$. If two messages $m_1$ and $m_2$ have the same timestamp, we break ties using their message identifier. More precisely, if $m_1.ts = m_2.ts \wedge m_1.id < m_2.id$, we consider that $m_1.ts < m_2.ts$.

%The consensus instances within $g$ are decided in the same order for every correct process of $g$. Such order is used to define the delivery order of $g$'s messages as well. We want this order to be consistent with the timestamps of the messages, so such timestamps can be changed if they are not in accordance with the delivery order.

@@

%To allow for the timestamp of messages to be increased and still have these messages delivered as soon as possible, their timestamps are increased by an infinitesimal value. For that reason, each timestamp value will consist of a real-time clock value, and a sequence value, which is used only when messages need to have their timestamps changed. Therefore, we have that $m.ts = (rtc, seq)$, where $rtc$ is some value related to real-time clock of a process and $seq$ is a sequence number to define an order between messages with the same $rtc$. Then we have:

%\begin{align*}
%m.ts < m'.ts \Longleftrightarrow m.ts.rtc < m'.ts.rtc \vee (m.ts.rtc = m'.ts.rtc \wedge m.ts.seq < %m'.ts.seq)\\
%%&\vee (m.ts.rtc = m'.ts.rtc \wedge m.ts.seq = m'.ts.seq \wedge m.ts.p_{id} < m'.ts.p_{id})
%\end{align*}

@@

@@

\blankline
\INDENT{\textbf{when} inserting a message $m$ into $\decided$}
  \INDENT{\textbf{for all} $h \in m.dst \cup \{g\}$ \textbf{do}}
    \STATE $lastBarrierCreated(h) \leftarrow m.ts$
  \ENDINDENT  
\ENDINDENT

\blankline
\INDENT{\textbf{when} $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow lastBarrierCreated(h) + barrierThreshold$ \label{algline:samenullid}
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $\localmsgs \leftarrow \localmsgs \cup \{null\}$ \label{algline:enqnull}
  \STATE $lastBarrierCreated(h) \leftarrow null.ts$
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

The problem with addressing liveness this way is that a process $q$ from group $h$ might have received some message $m$ and has also just received a barrier $b$ from $g \in sendersTo(h)$, such that $b < m.ts$. This would mean that, if $g$ has no messages to send to $h$, $q$ will have to wait for, at least, $barrierThreshold$ -- maybe just to receive from $g$ some barrier $c : c < m.ts$, having to wait again and so on. How long exactly it will take for $q$ to finally \amdel{} $m$ depends on several variables: besides the value of $barrierThreshold$, it dependon on how far in the past $m$ was created and how long it takes for some barrier $d: d > m.ts$ to arrive at $q$.% Because of that, the time between \amcast{}ing and \cons{}ing $m$ be, in the worst case, $2w(p) + 4\tconsm + barrierThreshold$, where $w(p)$ is the wait window of the process $p \in G$, and \tcons\ is counted twice because there are two consensus proposals -- first in $G'$ regarding $m$, and then in $G$ regarding the empty message.

It is necessary to guarantee that, whenever the $barrierThreshold$ time has elapsed since the last message to a given group $h$ has been created, an empty message will eventually be decided and sent to $g$, so that progression in $g$ is guaranteed. Therefore, such empty message is created and proposed by every process in each group $g$: if some process of $g$ does not propose the empty message, it might never be decided in such group. %Besides, they are excluded from the $\localmsgs$ set only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such set and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
%To prevent the multiple $null$ messages created simultaneously by different processes within each group from being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent. However, deciding and delivering them would not make the algorithm incorrect.%\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

\section{Optimized atomic multicast}
\label{sec:optimized}

\subsection{Requesting empty messages}
\label{sec:nullondemand}

@@

 \subsubsection{Performance analysis}

@@

\subsection{Optimistic Delivery}
\label{sec:optdel}
